---
title: "Module 3: Circle Secrets"
subtitle: "The inscribed angle"
output:
  xaringan::moon_reader:
    css: [default, default-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = "center", dev = "png",
                      fig.width = 5, fig.height = 5, dpi = 150)
blue   <- "#1769aa"
orange <- "#c2410c"
```

class: center, middle

# Circle Secrets

### The inscribed angle

Draw a circle, a diameter straight across it, and pick any point on the rim. Join
it to the two ends of the diameter. The angle you get is a right angle. Always.

---

```{css, echo=FALSE}
.why { clear: both; margin-top: 0.1em; font-size: 85%; }
```

# The setup

Everything today lives on a single circle, with two ways to "see" the same pair
of points.

--

- **Center angle:** the angle at the center **O**, looking out at two rim points
  **A** and **B**.

--

- **Rim angle (the inscribed angle):** the angle at a point **P** *on* the
  circle, looking at those same two points A and B.

--

One circle, two angles on the same pair of points. The whole module is one fact
about how those two angles are related.

---

# The hook: Thales

**Stand at any point on a circle's rim and look at the two ends of a diameter.
What angle do you see there, and does moving the point change it?**

--

.pull-left[
1. Draw a diameter **AB**.
2. Pick any point **P** on the circle and join it to A and B.
3. Measure the angle **APB**.
4. Now slide **P** somewhere else and measure again.
]

--

.pull-right[
```{r thales, fig.width=5, fig.height=3.5}
par(mar=c(0,0,0,0), pty="s")
t <- seq(0, 2*pi, length.out=300)
plot(cos(t), sin(t), type="l", col="grey70", asp=1, axes=FALSE, xlab="", ylab="",
     xlim=c(-1.28,1.28), ylim=c(-1.28,1.28))
A <- c(-1,0); B <- c(1,0); O <- c(0,0)
segments(A[1],A[2], B[1],B[2], col=orange, lwd=2)
angs <- c(55, 100, 140) * pi/180
for (a in angs) {
  P <- c(cos(a), sin(a))
  segments(P[1],P[2], A[1],A[2], col=blue, lwd=2)
  segments(P[1],P[2], B[1],B[2], col=blue, lwd=2)
  points(P[1], P[2], pch=19, cex=1.1)
}
points(rbind(A,B), pch=19, cex=1.3)
points(0, 0, pch=3, cex=1.2)
text(A[1]-0.13, A[2]-0.04, "A", cex=1.4)
text(B[1]+0.13, B[2]-0.04, "B", cex=1.4)
text(0.07, -0.13, "O", cex=1.1)
text(cos(100*pi/180)-0.04, sin(100*pi/180)+0.13, "P", cex=1.3)
```
]

--

.why[
It is **90 degrees**, every time: high, low, near the edge. The angle "standing on
a diameter" is *always* a right angle. Why can it never be 88 or 93?
]

---

# The bigger secret

**Forget the diameter. Put A and B *anywhere* on the circle. The center sees them
at one angle and a rim point sees them at another. How are those two angles
related?**

--

.pull-left[
1. Put **A** and **B** anywhere on the circle.
2. Draw the **center angle AOB**.
3. Pick **P** on the longer arc and draw the **rim angle APB**.
4. Compare the two.
]

--

.pull-right[
```{r rim, fig.width=5, fig.height=3.5}
par(mar=c(0,0,0,0), pty="s")
d2r <- function(d) d * pi/180
t <- seq(0, 2*pi, length.out=300)
plot(cos(t), sin(t), type="l", col="grey70", asp=1, axes=FALSE, xlab="", ylab="",
     xlim=c(-1.28,1.28), ylim=c(-1.28,1.28))
A <- c(cos(d2r(200)), sin(d2r(200)))
B <- c(cos(d2r(340)), sin(d2r(340)))
P <- c(cos(d2r(90)),  sin(d2r(90)))
O <- c(0,0)
segments(O[1],O[2], A[1],A[2], col=orange, lwd=2)
segments(O[1],O[2], B[1],B[2], col=orange, lwd=2)
segments(P[1],P[2], A[1],A[2], col=blue, lwd=2)
segments(P[1],P[2], B[1],B[2], col=blue, lwd=2)
points(rbind(A,B,P,O), pch=19, cex=1.2)
text(A[1]-0.12, A[2], "A", cex=1.3)
text(B[1]+0.12, B[2], "B", cex=1.3)
text(P[1], P[2]+0.13, "P", cex=1.3)
text(O[1]-0.1, O[2]+0.12, "O", cex=1.1)
text(0, 0.78, "t",  col=blue,   cex=1.3)
text(0, -0.2, "2t", col=orange, cex=1.2)
```
]

--

.why[
The rim angle is exactly **half** the center angle. If the center angle is **2t**,
the rim angle is **t**. The diameter was just the special case where the center
angle was a straight 180.
]

---

# Why must it be true?

**You measured it and it held. But measuring is not proving. What forces the rim
angle to be exactly half, for every position of P?**

--

.pull-left[
Take the clean case: the line **PO** runs through the center to the far point
**B**, so **PB** is a diameter. Draw the radius **OA**.

- **OA** and **OP** are both radii, so triangle **OAP** is **isosceles**.
- Equal sides face equal angles, so angle OPA = angle OAP. Call each **t**.
- Those three angles of triangle OAP add to 180 (you proved this in Module 2), so
  angle AOP = 180 - 2t.
- Angles AOP and AOB share the straight line PB, so they add to 180 too. Hence
  angle AOB = 180 - (180 - 2t) = **2t**.
]

--

.pull-right[
```{r proof, fig.width=5, fig.height=3.4}
par(mar=c(0,0,0,0), pty="s")
d2r <- function(d) d * pi/180
t <- seq(0, 2*pi, length.out=300)
plot(cos(t), sin(t), type="l", col="grey70", asp=1, axes=FALSE, xlab="", ylab="",
     xlim=c(-1.28,1.28), ylim=c(-1.28,1.28))
P <- c(0,1); B <- c(0,-1); O <- c(0,0)
A <- c(cos(d2r(210)), sin(d2r(210)))
segments(P[1],P[2], B[1],B[2], col="grey50", lwd=1.6)
segments(O[1],O[2], A[1],A[2], col=orange, lwd=2)
segments(P[1],P[2], A[1],A[2], col=blue, lwd=2)
points(rbind(A,B,P,O), pch=19, cex=1.2)
text(P[1], P[2]+0.13, "P", cex=1.3)
text(B[1], B[2]-0.13, "B", cex=1.3)
text(A[1]-0.13, A[2], "A", cex=1.3)
text(O[1]+0.12, O[2]+0.1, "O", cex=1.1)
text(-0.13, 0.58, "t",  col=blue,   cex=1.2)
text(-0.58, -0.32, "t", col=orange, cex=1.2)
text(-0.16, -0.26, "2t", col=orange, cex=1.1)
```
]

--

.why[
Since O sits on PB, the rim angle APB is just angle OPA = **t**. Center angle =
**2t**. The rim is half the center.
]

---

# Thales falls out for free

**We just proved the rim angle is half the center angle. So what does that force
Thales's right angle to be?**

--

When A and B are the ends of a **diameter**, the center angle AOB is a straight
line: **180 degrees**.

--

Half of 180 is **90**.

--

So Thales is not a separate miracle. It is the inscribed angle theorem with the
center angle stretched flat. One theorem, both facts.

---

# Slide the rim point along the arc

**Move P to a new spot Q on the same arc and look at AB again. Does the angle it
sees AB at get bigger, get smaller, or stay put?**

--

.pull-left[
1. Keep A and B fixed; the center O never moves.
2. So the center angle AOB is one fixed size.
3. Any rim point on the arc sees AB as an inscribed angle, half of that fixed
   center angle.
]

--

.pull-right[
```{r samearc, fig.width=5, fig.height=3.5}
par(mar=c(0,0,0,0), pty="s")
d2r <- function(d) d * pi/180
t <- seq(0, 2*pi, length.out=300)
plot(cos(t), sin(t), type="l", col="grey70", asp=1, axes=FALSE, xlab="", ylab="",
     xlim=c(-1.28,1.28), ylim=c(-1.28,1.28))
A <- c(cos(d2r(200)), sin(d2r(200)))
B <- c(cos(d2r(340)), sin(d2r(340)))
P <- c(cos(d2r(115)), sin(d2r(115)))
Q <- c(cos(d2r(55)),  sin(d2r(55)))
for (X in list(P, Q)) {
  segments(X[1],X[2], A[1],A[2], col=blue, lwd=2)
  segments(X[1],X[2], B[1],B[2], col=blue, lwd=2)
}
points(rbind(A,B,P,Q), pch=19, cex=1.2)
text(A[1]-0.12, A[2], "A", cex=1.3)
text(B[1]+0.12, B[2], "B", cex=1.3)
text(P[1]-0.06, P[2]+0.13, "P", cex=1.3)
text(Q[1]+0.1,  Q[2]+0.1,  "Q", cex=1.3)
```
]

--

.why[
Half of one fixed thing is one fixed thing, so **every** point on the arc sees AB
at the **same** angle: APB = AQB, wherever P and Q sit on the arc.
]

---

# Push further: cyclic quadrilaterals

**Put four points on a circle and join them in order. Pick two opposite corners
and add their angles. What do you always get?**

--

.pull-left[
- Each corner is an inscribed angle, half the arc it stands on.
- Two opposite corners stand on the two arcs that together cover the whole
  circle.
]

--

.pull-right[
```{r cyclic, fig.width=5, fig.height=3.5}
par(mar=c(0,0,0,0), pty="s")
d2r <- function(d) d * pi/180
t <- seq(0, 2*pi, length.out=300)
plot(cos(t), sin(t), type="l", col="grey70", asp=1, axes=FALSE, xlab="", ylab="",
     xlim=c(-1.28,1.28), ylim=c(-1.28,1.28))
ang <- c(80, 165, 250, 335)
V <- cbind(cos(d2r(ang)), sin(d2r(ang)))
polygon(V[,1], V[,2], border=blue, lwd=3)
points(V, pch=19, cex=1.2)
labs <- c("A","B","C","D")
for (i in 1:4) text(V[i,1]*1.16, V[i,2]*1.16, labs[i], cex=1.3)
```
]

--

.why[
**Opposite angles add to 180.** The two arcs cover the whole circle (360), and
half of 360 is 180. "These four points lie on one circle" is one of the most
powerful things to spot in a contest problem, and this is exactly why.
]

---

# Competition gem: crossing chords

.pull-left[
Two chords **AB** and **CD** cross at a point **X** inside the circle.

--

Triangle **AXC** and triangle **DXB** have the same shape (similar). Hunt for two
pairs of equal inscribed angles standing on the same arcs.

--

The payoff is the classic length rule:

`AX times XB = CX times XD`
]

.pull-right[
```{r chords, fig.width=5, fig.height=4.6}
par(mar=c(0,0,0,0), pty="s")
d2r <- function(d) d * pi/180
t <- seq(0, 2*pi, length.out=300)
plot(cos(t), sin(t), type="l", col="grey70", asp=1, axes=FALSE, xlab="", ylab="",
     xlim=c(-1.28,1.28), ylim=c(-1.28,1.28))
A <- c(cos(d2r(150)), sin(d2r(150)))
B <- c(cos(d2r(350)), sin(d2r(350)))
C <- c(cos(d2r(80)),  sin(d2r(80)))
D <- c(cos(d2r(250)), sin(d2r(250)))
inter <- function(A,B,C,D){
  d1 <- B - A; d2 <- D - C
  M <- matrix(c(d1[1], -d2[1], d1[2], -d2[2]), nrow=2, byrow=TRUE)
  s <- solve(M, C - A)
  A + s[1] * d1
}
X <- inter(A,B,C,D)
segments(C[1],C[2], A[1],A[2], col="grey55", lwd=1, lty=2)
segments(D[1],D[2], B[1],B[2], col="grey55", lwd=1, lty=2)
segments(A[1],A[2], B[1],B[2], col=blue,   lwd=2)
segments(C[1],C[2], D[1],D[2], col=orange, lwd=2)
points(rbind(A,B,C,D), pch=19, cex=1.2)
points(X[1], X[2], pch=19, cex=1.2)
text(A[1]-0.12, A[2], "A", cex=1.3)
text(B[1]+0.12, B[2], "B", cex=1.3)
text(C[1], C[2]+0.13, "C", cex=1.3)
text(D[1], D[2]-0.13, "D", cex=1.3)
text(X[1]+0.1, X[2]+0.1, "X", cex=1.2)
```
]

---

class: center, middle

# Next up

**Module 4:** tilings. Why do only three regular shapes fill a floor with no gaps?
The answer is an angle that has to divide 360, and you already know how to chase
angles around a point.
