---
title: "Module 3: Circle Secrets"
subtitle: "The inscribed angle"
output:
  xaringan::moon_reader:
    css: [default, default-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = "center", dev = "png",
                      fig.width = 5, fig.height = 5, dpi = 150)
blue   <- "#1769aa"
orange <- "#c2410c"
```

class: center, middle

# Circle Secrets

### The inscribed angle

Draw a circle, a diameter straight across it, and pick any point on the rim. Join
it to the two ends of the diameter. The angle you get is a right angle. Always.

---

# The setup

Everything today lives on a single circle, with two ways to "see" the same pair
of points.

--

- **Center angle:** the angle at the center **O**, looking out at two rim points
  **A** and **B**.

--

- **Rim angle (the inscribed angle):** the angle at a point **P** *on* the
  circle, looking at those same two points A and B.

--

One circle, two angles on the same pair of points. The whole module is one fact
about how those two angles are related.

---

# The hook: Thales

.pull-left[
Draw a diameter **AB**. Pick any point **P** on the circle and join it to A and B.

--

Measure angle **APB**: it is **90 degrees**.

--

Move P anywhere else on the rim. Still 90. High, low, near the edge.

--

The angle "standing on a diameter" is *always* a right angle. Why can it never be
88 or 93?
]

.pull-right[
```{r thales, fig.width=5, fig.height=4.6}
par(mar=c(0,0,0,0), pty="s")
t <- seq(0, 2*pi, length.out=300)
plot(cos(t), sin(t), type="l", col="grey70", asp=1, axes=FALSE, xlab="", ylab="",
     xlim=c(-1.28,1.28), ylim=c(-1.28,1.28))
A <- c(-1,0); B <- c(1,0); O <- c(0,0)
segments(A[1],A[2], B[1],B[2], col=orange, lwd=2)
angs <- c(55, 100, 140) * pi/180
for (a in angs) {
  P <- c(cos(a), sin(a))
  segments(P[1],P[2], A[1],A[2], col=blue, lwd=2)
  segments(P[1],P[2], B[1],B[2], col=blue, lwd=2)
  points(P[1], P[2], pch=19, cex=1.1)
}
points(rbind(A,B), pch=19, cex=1.3)
points(0, 0, pch=3, cex=1.2)
text(A[1]-0.13, A[2]-0.04, "A", cex=1.4)
text(B[1]+0.13, B[2]-0.04, "B", cex=1.4)
text(0.07, -0.13, "O", cex=1.1)
text(cos(100*pi/180)-0.04, sin(100*pi/180)+0.13, "P", cex=1.3)
```
]

---

# The bigger secret

.pull-left[
Forget the diameter. Put **A** and **B** *anywhere* on the circle.

- Draw the **center angle AOB**.
- Pick **P** on the longer arc and draw the **rim angle APB**.

--

The rim angle is exactly **half** the center angle.

--

If the center angle is **2t**, the rim angle is **t**. The diameter was just the
special case where the center angle was a straight 180.
]

.pull-right[
```{r rim, fig.width=5, fig.height=4.6}
par(mar=c(0,0,0,0), pty="s")
d2r <- function(d) d * pi/180
t <- seq(0, 2*pi, length.out=300)
plot(cos(t), sin(t), type="l", col="grey70", asp=1, axes=FALSE, xlab="", ylab="",
     xlim=c(-1.28,1.28), ylim=c(-1.28,1.28))
A <- c(cos(d2r(200)), sin(d2r(200)))
B <- c(cos(d2r(340)), sin(d2r(340)))
P <- c(cos(d2r(90)),  sin(d2r(90)))
O <- c(0,0)
segments(O[1],O[2], A[1],A[2], col=orange, lwd=2)
segments(O[1],O[2], B[1],B[2], col=orange, lwd=2)
segments(P[1],P[2], A[1],A[2], col=blue, lwd=2)
segments(P[1],P[2], B[1],B[2], col=blue, lwd=2)
points(rbind(A,B,P,O), pch=19, cex=1.2)
text(A[1]-0.12, A[2], "A", cex=1.3)
text(B[1]+0.12, B[2], "B", cex=1.3)
text(P[1], P[2]+0.13, "P", cex=1.3)
text(O[1]-0.1, O[2]+0.12, "O", cex=1.1)
text(0, 0.78, "t",  col=blue,   cex=1.3)
text(0, -0.2, "2t", col=orange, cex=1.2)
```
]

---

# Why must it be true?

.pull-left[
Take the clean case: the line **PO** runs through the center to the far point
**B**, so **PB** is a diameter. Draw the radius **OA**.

- **OA** and **OP** are both radii, so triangle **OAP** is **isosceles**.
- Equal sides face equal angles, so angle OPA = angle OAP. Call each **t**.
- The center angle **AOB** is the *exterior* angle of the triangle at O, so it
  equals t + t = **2t**.

--

Rim angle = **t**. Center angle = **2t**. The rim is half the center.
]

.pull-right[
```{r proof, fig.width=5, fig.height=4.6}
par(mar=c(0,0,0,0), pty="s")
d2r <- function(d) d * pi/180
t <- seq(0, 2*pi, length.out=300)
plot(cos(t), sin(t), type="l", col="grey70", asp=1, axes=FALSE, xlab="", ylab="",
     xlim=c(-1.28,1.28), ylim=c(-1.28,1.28))
P <- c(0,1); B <- c(0,-1); O <- c(0,0)
A <- c(cos(d2r(210)), sin(d2r(210)))
segments(P[1],P[2], B[1],B[2], col="grey50", lwd=1.6)
segments(O[1],O[2], A[1],A[2], col=orange, lwd=2)
segments(P[1],P[2], A[1],A[2], col=blue, lwd=2)
points(rbind(A,B,P,O), pch=19, cex=1.2)
text(P[1], P[2]+0.13, "P", cex=1.3)
text(B[1], B[2]-0.13, "B", cex=1.3)
text(A[1]-0.13, A[2], "A", cex=1.3)
text(O[1]+0.12, O[2]+0.1, "O", cex=1.1)
text(-0.13, 0.58, "t",  col=blue,   cex=1.2)
text(-0.58, -0.32, "t", col=orange, cex=1.2)
text(-0.16, -0.26, "2t", col=orange, cex=1.1)
```
]

---

# Thales falls out for free

When A and B are the ends of a **diameter**, the center angle AOB is a straight
line: **180 degrees**.

--

Half of 180 is **90**.

--

So Thales is not a separate miracle. It is the inscribed angle theorem with the
center angle stretched flat. One theorem, both facts.

---

# Same arc, same angle

.pull-left[
Every rim point on the same arc sees the *same* center angle AOB. The center O
never moves.

--

Half of one fixed thing is one fixed thing.

--

So **every** point on that arc sees AB at the **same** angle: APB = AQB, no matter
where P and Q sit on the arc.
]

.pull-right[
```{r samearc, fig.width=5, fig.height=4.6}
par(mar=c(0,0,0,0), pty="s")
d2r <- function(d) d * pi/180
t <- seq(0, 2*pi, length.out=300)
plot(cos(t), sin(t), type="l", col="grey70", asp=1, axes=FALSE, xlab="", ylab="",
     xlim=c(-1.28,1.28), ylim=c(-1.28,1.28))
A <- c(cos(d2r(200)), sin(d2r(200)))
B <- c(cos(d2r(340)), sin(d2r(340)))
P <- c(cos(d2r(115)), sin(d2r(115)))
Q <- c(cos(d2r(55)),  sin(d2r(55)))
for (X in list(P, Q)) {
  segments(X[1],X[2], A[1],A[2], col=blue, lwd=2)
  segments(X[1],X[2], B[1],B[2], col=blue, lwd=2)
}
points(rbind(A,B,P,Q), pch=19, cex=1.2)
text(A[1]-0.12, A[2], "A", cex=1.3)
text(B[1]+0.12, B[2], "B", cex=1.3)
text(P[1]-0.06, P[2]+0.13, "P", cex=1.3)
text(Q[1]+0.1,  Q[2]+0.1,  "Q", cex=1.3)
```
]

---

# Push further: cyclic quadrilaterals

.pull-left[
Put four points on a circle and join them into a quadrilateral.

--

**Opposite angles add to 180.**

Each corner is an inscribed angle. The arcs they stand on pair up to cover the
whole circle, and the whole circle is 360. Half of 360 is 180.

--

"These four points lie on one circle" is one of the most powerful things to spot
in a contest problem, and this is exactly why.
]

.pull-right[
```{r cyclic, fig.width=5, fig.height=4.6}
par(mar=c(0,0,0,0), pty="s")
d2r <- function(d) d * pi/180
t <- seq(0, 2*pi, length.out=300)
plot(cos(t), sin(t), type="l", col="grey70", asp=1, axes=FALSE, xlab="", ylab="",
     xlim=c(-1.28,1.28), ylim=c(-1.28,1.28))
ang <- c(80, 165, 250, 335)
V <- cbind(cos(d2r(ang)), sin(d2r(ang)))
polygon(V[,1], V[,2], border=blue, lwd=3)
points(V, pch=19, cex=1.2)
labs <- c("A","B","C","D")
for (i in 1:4) text(V[i,1]*1.16, V[i,2]*1.16, labs[i], cex=1.3)
```
]

---

# Competition gem: crossing chords

.pull-left[
Two chords **AB** and **CD** cross at a point **X** inside the circle.

--

Triangle **AXC** and triangle **DXB** have the same shape (similar). Hunt for two
pairs of equal inscribed angles standing on the same arcs.

--

The payoff is the classic length rule:

`AX times XB = CX times XD`
]

.pull-right[
```{r chords, fig.width=5, fig.height=4.6}
par(mar=c(0,0,0,0), pty="s")
d2r <- function(d) d * pi/180
t <- seq(0, 2*pi, length.out=300)
plot(cos(t), sin(t), type="l", col="grey70", asp=1, axes=FALSE, xlab="", ylab="",
     xlim=c(-1.28,1.28), ylim=c(-1.28,1.28))
A <- c(cos(d2r(150)), sin(d2r(150)))
B <- c(cos(d2r(350)), sin(d2r(350)))
C <- c(cos(d2r(80)),  sin(d2r(80)))
D <- c(cos(d2r(250)), sin(d2r(250)))
inter <- function(A,B,C,D){
  d1 <- B - A; d2 <- D - C
  M <- matrix(c(d1[1], -d2[1], d1[2], -d2[2]), nrow=2, byrow=TRUE)
  s <- solve(M, C - A)
  A + s[1] * d1
}
X <- inter(A,B,C,D)
segments(C[1],C[2], A[1],A[2], col="grey55", lwd=1, lty=2)
segments(D[1],D[2], B[1],B[2], col="grey55", lwd=1, lty=2)
segments(A[1],A[2], B[1],B[2], col=blue,   lwd=2)
segments(C[1],C[2], D[1],D[2], col=orange, lwd=2)
points(rbind(A,B,C,D), pch=19, cex=1.2)
points(X[1], X[2], pch=19, cex=1.2)
text(A[1]-0.12, A[2], "A", cex=1.3)
text(B[1]+0.12, B[2], "B", cex=1.3)
text(C[1], C[2]+0.13, "C", cex=1.3)
text(D[1], D[2]-0.13, "D", cex=1.3)
text(X[1]+0.1, X[2]+0.1, "X", cex=1.2)
```
]

---

class: center, middle

# Next up

**Module 4:** tilings. Why do only three regular shapes fill a floor with no gaps?
The answer is an angle that has to divide 360, and you already know how to chase
angles around a point.
