---
title: "Module 1: Compass and Straightedge"
subtitle: "The oldest game in math"
output:
  xaringan::moon_reader:
    css: [default, default-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = "center", dev = "png",
                      fig.width = 5, fig.height = 5, dpi = 150)
blue   <- "#1769aa"
orange <- "#c2410c"
red_c  <- "#c0392b"
grn_c  <- "#1b7a3d"

# Perpendicular-bisector figure. step 0 = plain construction; steps 1-3 add the
# colored highlights that match the proof bullets on the build slides.
draw_bisect <- function(step = 0) {
  par(mar=c(0,0,0,0), pty="s")
  A <- c(0,0); B <- c(2,0); r <- 1.3; M <- c(1,0)
  yc <- sqrt(r^2 - 1); P <- c(1, yc); Q <- c(1, -yc)
  plot(NA, xlim=c(-0.55,2.55), ylim=c(-1.3,1.4), asp=1, axes=FALSE, xlab="", ylab="")
  aA <- seq(-52, 52, length.out=120) * pi/180
  lines(A[1]+r*cos(aA), A[2]+r*sin(aA), col="grey75", lwd=2)
  aB <- seq(128, 232, length.out=120) * pi/180
  lines(B[1]+r*cos(aB), B[2]+r*sin(aB), col="grey75", lwd=2)
  segments(A[1],A[2], B[1],B[2], col="grey30", lwd=3)
  segments(M[1], -1.05, M[1], 1.1, col=blue, lwd=4)
  sq <- 0.13
  polygon(c(1,1+sq,1+sq,1), c(0,0,sq,sq), border=orange, lwd=2)
  segments(0.5,-0.07,0.5,0.07, col=orange, lwd=2)
  segments(1.5,-0.07,1.5,0.07, col=orange, lwd=2)
  if (step == 1) {
    for (X in list(A,B)) segments(P[1],P[2],X[1],X[2], col=red_c, lwd=3)
    for (X in list(A,B)) segments(Q[1],Q[2],X[1],X[2], col=red_c, lwd=3)
  }
  if (step == 2) {
    polygon(c(P[1],A[1],Q[1]), c(P[2],A[2],Q[2]), border=red_c, lwd=3, col=adjustcolor(red_c,0.10))
    polygon(c(P[1],B[1],Q[1]), c(P[2],B[2],Q[2]), border=grn_c, lwd=3, col=adjustcolor(grn_c,0.10))
  }
  if (step == 3) {
    polygon(c(A[1],P[1],M[1]), c(A[2],P[2],M[2]), border=red_c, lwd=3, col=adjustcolor(red_c,0.12))
    polygon(c(B[1],P[1],M[1]), c(B[2],P[2],M[2]), border=grn_c, lwd=3, col=adjustcolor(grn_c,0.12))
  }
  points(rbind(A,B), pch=19, cex=1.3)
  points(rbind(P,Q), pch=19, cex=1.0, col="grey30")
  points(M[1],M[2], pch=19, cex=1.0, col=blue)
  text(A[1]-0.14, -0.02, "A", cex=1.4)
  text(B[1]+0.14, -0.02, "B", cex=1.4)
  text(0.82, -0.16, "M", cex=1.2, col=blue)
  text(1.17, yc+0.03, "P", cex=1.2, col="grey30")
  text(1.17, -yc-0.03, "Q", cex=1.2, col="grey30")
}

# Hinge illustration for the "two sides + the angle between them" rule.
# Solid = the fixed triangle; dashed ghost = a different hinge angle, different triangle.
draw_hinge <- function() {
  par(mar=c(0,0,0,0), pty="s")
  P <- c(0,0); La <- 1.7; Lm <- 1.25
  A <- P + c(La, 0)
  th1 <- 52*pi/180; M  <- P + Lm*c(cos(th1), sin(th1))
  th2 <- 88*pi/180; M2 <- P + Lm*c(cos(th2), sin(th2))
  plot(NA, xlim=c(-0.15,1.95), ylim=c(-0.35,1.5), asp=1, axes=FALSE, xlab="", ylab="")
  segments(P[1],P[2], M2[1],M2[2], col=adjustcolor(grn_c,0.35), lwd=3, lty=2)
  segments(A[1],A[2], M2[1],M2[2], col="grey65", lwd=2, lty=2)
  points(M2[1],M2[2], pch=1, cex=1.1, col=adjustcolor(grn_c,0.6))
  text(M2[1]-0.14, M2[2]+0.06, "M?", cex=1.0, col=adjustcolor(grn_c,0.75))
  segments(P[1],P[2], A[1],A[2], col=red_c, lwd=4)
  segments(P[1],P[2], M[1],M[2], col=grn_c, lwd=4)
  segments(A[1],A[2], M[1],M[2], col="grey40", lwd=3)
  aa <- seq(0, th1, length.out=40); rr <- 0.34
  lines(P[1]+rr*cos(aa), P[2]+rr*sin(aa), col="grey30", lwd=2)
  points(rbind(P,A,M), pch=19, cex=1.1)
  text(P[1]-0.02, P[2]-0.15, "P", cex=1.3)
  text(A[1]+0.08, A[2]-0.02, "A", cex=1.3)
  text(M[1]+0.07, M[2]+0.07, "M", cex=1.3)
}

# Hexaflexagon build figures (bonus tie-in to the Move 3 hexagon).
draw_flex_strip <- function(n = 10) {
  par(mar=c(0,0,0,0), pty="s")
  s <- 1; h <- sqrt(3)/2
  xmax <- (n+1)*s/2
  plot(NA, xlim=c(-0.15, xmax+0.15), ylim=c(-0.2, h+0.2), asp=1, axes=FALSE, xlab="", ylab="")
  for (i in 1:n) {
    if (i %% 2 == 1) {
      x <- c((i-1)*s/2, (i+1)*s/2, i*s/2); y <- c(0, 0, h)
    } else {
      x <- c((i-1)*s/2, (i+1)*s/2, i*s/2); y <- c(h, h, 0)
    }
    col_i <- if (i %% 2 == 1) adjustcolor(blue, 0.15) else adjustcolor(orange, 0.15)
    polygon(x, y, border="grey30", lwd=1.5, col=col_i)
    text(mean(x), mean(y), i, cex=0.75, col="grey30")
  }
}

draw_flex_ring <- function() {
  par(mar=c(0,0,0,0), pty="s")
  ang <- seq(0, 2*pi, length.out=7)
  x <- cos(ang); y <- sin(ang)
  plot(NA, xlim=c(-1.3,1.3), ylim=c(-1.3,1.3), asp=1, axes=FALSE, xlab="", ylab="")
  polygon(x, y, border=blue, lwd=4)
  segments(0,0, x[-7], y[-7], col=orange, lwd=2)
  points(x[-7], y[-7], pch=19, cex=1.2)
  points(0,0, pch=3, cex=1.4)
  polygon(c(0, x[6], x[7]), c(0, y[6], y[7]), border=red_c, lwd=3, col=adjustcolor(red_c, 0.15))
  text(mean(c(0,x[6],x[7])), mean(c(0,y[6],y[7])), "glue", cex=0.85, col=red_c)
}

draw_flex_pinch <- function() {
  par(mar=c(0,0,0,0), pty="s")
  ang <- seq(0, 2*pi, length.out=7)
  x <- cos(ang); y <- sin(ang)
  plot(NA, xlim=c(-1.3,1.3), ylim=c(-1.3,1.3), asp=1, axes=FALSE, xlab="", ylab="")
  polygon(x, y, border="grey50", lwd=3)
  for (i in c(1,3,5)) {
    arrows(x[i]*0.92, y[i]*0.92, x[i]*0.35, y[i]*0.35, col=red_c, lwd=3, length=0.12)
  }
  points(x[-7], y[-7], pch=19, cex=1.1, col="grey40")
}

draw_flex_newface <- function() {
  par(mar=c(0,0,0,0), pty="s")
  ang <- seq(0, 2*pi, length.out=7)
  x <- cos(ang); y <- sin(ang)
  plot(NA, xlim=c(-1.25,1.25), ylim=c(-1.25,1.25), asp=1, axes=FALSE, xlab="", ylab="")
  polygon(x, y, border=grn_c, lwd=4, col=adjustcolor(grn_c, 0.15))
  segments(0,0, x[-7], y[-7], col=grn_c, lwd=2)
  points(x[-7], y[-7], pch=19, cex=1.2, col=grn_c)
  points(0,0, pch=3, cex=1.4, col=grn_c)
}
```

class: center, middle

# Compass and Straightedge

### The oldest game in math

With one compass and one unmarked ruler, the ancient Greeks built perfect shapes
more than 2000 years ago. No measuring.

---

```{css, echo=FALSE}
.why { clear: both; margin-top: 0.1em; font-size: 85%; }
.red { color: #c0392b; font-weight: bold; }
.grn { color: #1b7a3d; font-weight: bold; }
```

# The whole game has two moves

You may only ever do two things:

--

1. **Straightedge:** draw a straight line through two points.

--

2. **Compass:** draw a circle, given a center and a radius you set by placing the
   two legs on two points.

--

That is it. Everything today is built from just these two moves. The ruler has
numbers, but we never use them. It is only an edge.

---

# The hook

Can you make a **perfect** equilateral triangle, where you are sure all three
sides are exactly equal...

--

...without measuring a single thing?

--

Not "looks equal." *Forced* to be equal. Let's build it, then see why it must be.

---

# Build it (this is Euclid, Proposition 1)

**How do you *force* three equal sides, not just eyeball them? Two circles do it.**

--

.pull-left[
1. Draw a segment. Call the ends **A** and **B**.
2. Compass point on **A**, pencil on **B**: draw a circle.
3. Same width, point on **B**, pencil on **A**: draw a second circle.
4. The circles cross. Call the top crossing **C**.
5. Draw **AC** and **BC**.
]

--

.pull-right[
```{r tri, fig.width=5, fig.height=4.2}
par(mar=c(0,0,0,0), pty="s")
A <- c(0,0); B <- c(1,0); r <- 1
t <- seq(0,2*pi,length.out=300)
plot(NA, xlim=c(-1.15,2.15), ylim=c(-1.25,1.35), asp=1, axes=FALSE, xlab="", ylab="")
lines(A[1]+r*cos(t), A[2]+r*sin(t), col="grey70")
lines(B[1]+r*cos(t), B[2]+r*sin(t), col="grey70")
C <- c(0.5, sqrt(3)/2)
polygon(c(A[1],B[1],C[1]), c(A[2],B[2],C[2]), border=blue, lwd=4)
points(rbind(A,B,C), pch=19, cex=1.3)
text(A[1]-0.1,A[2]-0.05,"A", cex=1.4)
text(B[1]+0.1,B[2]-0.05,"B", cex=1.4)
text(C[1],C[2]+0.13,"C", cex=1.4)
```
]

---

# Why must the sides be equal?

**You never measured. So how do we *know* all three came out equal?**

--

- **AB** is the radius of both circles (you never changed the width).
- **AC** is a radius of the circle centered at A, so `AC = AB`.
- **BC** is a radius of the circle centered at B, so `BC = AB`.

--

All three equal the same radius, so they equal each other.

--

This is not "close." It is **forced**, for *any* starting segment AB. That jump,
from one drawing to every possible drawing, is what a **proof** is.

---

# Move 2: cut a segment exactly in half

**Can you find the exact middle of a segment, no ruler markings, and be *sure*
it is exact?**

--

.pull-left[
Same trick as the triangle, done twice:

1. Compass point on **A**, opened past halfway: arc above and below.
2. Same width, point on **B**: arcs crossing the first two.
3. Draw the line through the two crossings.
]

--

.pull-right[
```{r bisect0, fig.width=5, fig.height=3.8}
draw_bisect(0)
```
]

---

# Move 2: cut a segment exactly in half

**Why is it exact? No measuring happened, so we prove it.**

.pull-left[
- The red segments .red[PA], .red[PB], .red[QA], .red[QB] are each exactly one
  **radius** long: that is how the two arcs were drawn.
]

.pull-right[
```{r bisect1, fig.width=5, fig.height=3.8}
draw_bisect(1)
```
]

---

# Move 2: cut a segment exactly in half

**Why is it exact? No measuring happened, so we prove it.**

.pull-left[
- The red segments PA, PB, QA, QB are each one **radius** long.
- So triangle .red[PAQ] and triangle .grn[PBQ] have the same three sides, which
  forces them to be exact copies (mirror images). That makes **PQ** cut the angle
  at **P** (the angle **APB**, between **PA** and **PB**) into two equal halves.
]

.pull-right[
```{r bisect2, fig.width=5, fig.height=3.8}
draw_bisect(2)
```
]

---

# Move 2: cut a segment exactly in half

**A tool we need: why do two sides plus the angle between them fix a triangle?**

.pull-left[
Think of the two sides as sticks joined at a corner **P**, like a hinge.

- Leave the hinge loose and the far end swings (dashed): the third side gets
  longer or shorter, a *different* triangle each time.
- Fix the angle at the hinge and both far ends are locked. The third side has no
  choice left.

So two sides with a **fixed angle between them** allow exactly **one** triangle.
]

.pull-right[
```{r hinge, fig.width=5, fig.height=3.6}
draw_hinge()
```
]

---

# Move 2: cut a segment exactly in half

**Why is it exact? No measuring happened, so we prove it.**

.pull-left[
- The red segments PA, PB, QA, QB are each one **radius** long.
- Triangles PAQ and PBQ have the same three sides, so **PQ** halves the angle at **P**.
- Now triangle .red[APM] and triangle .grn[BPM] have two matching sides (shared
  **PM**, and **PA = PB**) with the **same angle at P** between them, so by the
  hinge rule they are exact copies.
- Exact copies have equal parts: **AM = BM**, so **M** is the midpoint. Their angles
  at **M** are equal too, and since **A, M, B** lie on a straight line those two
  angles add to 180, so each must be a **right angle**.
]

.pull-right[
```{r bisect3, fig.width=5, fig.height=3.8}
draw_bisect(3)
```
]

---

# Move 3: walk the compass around a circle

**Keep one compass width and step it around a circle, marking as you go. How many
marks until you land back on the start?**

--

.pull-left[
1. Draw a circle. **Do not change the compass width again.**
2. Put the point anywhere on the circle, mark where the pencil crosses it.
3. Walk to that mark, make the next one. Keep going.
4. Count the steps back to start.
]

--

.pull-right[
```{r hex, fig.width=5, fig.height=3.5}
par(mar=c(0,0,0,0), pty="s")
t <- seq(0,2*pi,length.out=300)
plot(cos(t), sin(t), type="l", col="grey70", asp=1, axes=FALSE, xlab="", ylab="")
ang <- seq(0, 2*pi, length.out=7)
x <- cos(ang); y <- sin(ang)
polygon(x, y, border=blue, lwd=4)
segments(0,0, x[-7], y[-7], col=orange, lwd=2)
points(x[-7], y[-7], pch=19, cex=1.3)
points(0,0, pch=3, cex=1.6)
```
]

--

.why[
**Exactly six, every time.** Connect the marks and you have a perfect hexagon.
]

---

# Why *exactly* six?

**Six every time, never five or seven. What forces the count?**

--

Connect the center to two next-door marks.

--

- Two of those sides are radii of the circle.
- The third side is your fixed compass width... which is also that radius.

--

So every little triangle is **equilateral**, which means a **60 degree** angle at
the center.

--

Angles all the way around the center add to **360**. And `360 / 60 = 6`.

The hexagon falls out of one surprising fact: a regular hexagon's side equals the
radius of the circle around it.

---

class: center, middle

# Bonus build: fold a hexaflexagon

You just built a hexagon with a compass and straightedge. Can you build the
same hexagon using nothing but folds?

Watch first: [Vi Hart, Hexaflexagons](https://www.khanacademy.org/math/math-for-fun-and-glory/vi-hart/hexaflexagons/v/hexaflexagons)

---

# Step 1: fold a chain of triangles

**Take a long strip of paper. Can you turn it into a chain of matching
triangles using only folds?**

--

.pull-left[
1. Fold one corner up so the bottom edge lines up with the side edge: that
   crease is exactly 60 degrees.
2. Fold the strip back the other way along the top edge of that triangle.
3. Keep folding back and forth, the same way, all the way down the strip.
4. Stop once you have **ten** triangles; trim off the rest.
]

--

.pull-right[
```{r flexstrip, fig.width=6, fig.height=2.6}
draw_flex_strip()
```
]

---

# Step 2: curl it into a ring

**Ten triangles in a row. What happens if you curl that strip around on
itself?**

--

.pull-left[
1. Using the same creases, curl the chain around into a flat loop.
2. Triangle 10 lands right behind triangle 1.
3. Glue or tape triangle 10 to the back of triangle 1.
]

--

.pull-right[
```{r flexring, fig.width=5, fig.height=4.2}
draw_flex_ring()
```
]

--

.why[
That loop is the same hexagon you built with the compass, just made of
folded paper triangles instead of arcs.
]

---

# Step 3: flex it open

**The hexagon looks finished. But it is hiding two more faces. How do you
get them out?**

--

.pull-left[
1. Pinch three corners that skip one another (1, 3, 5) and fold them up to
   meet in the middle.
2. The other three corners will pop open on their own.
3. Push down through the middle: a brand new face unfolds flat.
]

--

.pull-right[
```{r flexpinch, fig.width=5, fig.height=4}
draw_flex_pinch()
```
]

---

class: center, middle

# A new face!

```{r flexnew, fig.width=4.5, fig.height=4}
draw_flex_newface()
```

--

.why[
A trihexaflexagon like this one has **three** faces hiding in the same ring
of paper. Keep flexing (there is more than one way to fold it shut) to cycle
through all three. Color each face differently before you fold it, so you can
see them change.
]

---

# Push further

- **Bisect any angle** with the compass (hint: find two triangles that must be
  identical).
- **Build a square** using the perpendicular-bisector move twice.

--

- **The 2000-year cliffhanger.** Some shapes *can* be built and some *cannot*, and
  it took until the 1800s to prove which:
  - Pentagon (5 sides): **yes**. A 17-sided regular polygon: **yes** (Gauss, age 19).
  - 7-sided regular polygon: **no**. Trisecting a general angle: **no**. Squaring
    the circle: **no**.

These are real theorems, not "nobody figured it out yet."

---

# Competition gem

Someone hands you a circle but has **erased the center**. With only compass and
straightedge, find the center.

--

Hint: any chord's perpendicular bisector passes through the center. How many
chords do you need?

---

class: center, middle

# Next up

**Module 2:** tear the corners off any triangle. They always make a straight
line. Why *always*? That is where proof really begins.

Then play: **euclidea.xyz** turns these exact moves into puzzle levels.
