---
title: "Module 1: Compass and Straightedge"
subtitle: "The oldest game in math"
output:
  xaringan::moon_reader:
    css: [default, default-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = "center", dev = "png",
                      fig.width = 5, fig.height = 5, dpi = 150)
blue   <- "#1769aa"
orange <- "#c2410c"
```

class: center, middle

# Compass and Straightedge

### The oldest game in math

With one compass and one unmarked ruler, the ancient Greeks built perfect shapes
more than 2000 years ago. No measuring.

---

# The whole game has two moves

You may only ever do two things:

--

1. **Straightedge:** draw a straight line through two points.

--

2. **Compass:** draw a circle, given a center and a radius you set by placing the
   two legs on two points.

--

That is it. Everything today is built from just these two moves. The ruler has
numbers, but we never use them. It is only an edge.

---

# The hook

Can you make a **perfect** equilateral triangle, where you are sure all three
sides are exactly equal...

--

...without measuring a single thing?

--

Not "looks equal." *Forced* to be equal. Let's build it, then see why it must be.

---

# Build it (this is Euclid, Proposition 1)

.pull-left[
1. Draw a segment. Call the ends **A** and **B**.
2. Compass point on **A**, pencil on **B**: draw a circle.
3. Same width, point on **B**, pencil on **A**: draw a second circle.
4. The circles cross. Call the top crossing **C**.
5. Draw **AC** and **BC**.
]

.pull-right[
```{r tri, fig.width=5, fig.height=4.2}
par(mar=c(0,0,0,0), pty="s")
A <- c(0,0); B <- c(1,0); r <- 1
t <- seq(0,2*pi,length.out=300)
plot(NA, xlim=c(-1.15,2.15), ylim=c(-1.25,1.35), asp=1, axes=FALSE, xlab="", ylab="")
lines(A[1]+r*cos(t), A[2]+r*sin(t), col="grey70")
lines(B[1]+r*cos(t), B[2]+r*sin(t), col="grey70")
C <- c(0.5, sqrt(3)/2)
polygon(c(A[1],B[1],C[1]), c(A[2],B[2],C[2]), border=blue, lwd=4)
points(rbind(A,B,C), pch=19, cex=1.3)
text(A[1]-0.1,A[2]-0.05,"A", cex=1.4)
text(B[1]+0.1,B[2]-0.05,"B", cex=1.4)
text(C[1],C[2]+0.13,"C", cex=1.4)
```
]

---

# Why must the sides be equal?

You never measured. So how do we *know*?

--

- **AB** is the radius of both circles (you never changed the width).
- **AC** is a radius of the circle centered at A, so `AC = AB`.
- **BC** is a radius of the circle centered at B, so `BC = AB`.

--

All three equal the same radius, so they equal each other.

--

This is not "close." It is **forced**, for *any* starting segment AB. That jump,
from one drawing to every possible drawing, is what a **proof** is.

---

# Move 2: cut a segment exactly in half

The perpendicular bisector. Same trick, twice:

1. Compass point on **A**, opened past halfway, arc above and below.
2. Same width, point on **B**, arcs crossing the first two.
3. Draw the line through the two crossings.

--

That line cuts **AB** exactly in half **and** makes a perfect right angle, with no
protractor.

---

# Move 3: the six-petal flower

.pull-left[
1. Draw a circle. **Do not change the compass width again.**
2. Put the point anywhere on the circle, mark where the pencil crosses it.
3. Walk to that mark, make the next one. Keep going.
4. Count the steps back to start.

It is **exactly six**, every time. Connect them: a perfect hexagon.
]

.pull-right[
```{r hex, fig.width=5, fig.height=4.6}
par(mar=c(0,0,0,0), pty="s")
t <- seq(0,2*pi,length.out=300)
plot(cos(t), sin(t), type="l", col="grey70", asp=1, axes=FALSE, xlab="", ylab="")
ang <- seq(0, 2*pi, length.out=7)
x <- cos(ang); y <- sin(ang)
polygon(x, y, border=blue, lwd=4)
segments(0,0, x[-7], y[-7], col=orange, lwd=2)
points(x[-7], y[-7], pch=19, cex=1.3)
points(0,0, pch=3, cex=1.6)
```
]

---

# Why *exactly* six?

Connect the center to two next-door marks.

--

- Two of those sides are radii of the circle.
- The third side is your fixed compass width... which is also that radius.

--

So every little triangle is **equilateral**, which means a **60 degree** angle at
the center.

--

Angles all the way around the center add to **360**. And `360 / 60 = 6`.

The hexagon falls out of one surprising fact: a regular hexagon's side equals the
radius of the circle around it.

---

# Push further

- **Bisect any angle** with the compass (hint: find two triangles that must be
  identical).
- **Build a square** using the perpendicular-bisector move twice.

--

- **The 2000-year cliffhanger.** Some shapes *can* be built and some *cannot*, and
  it took until the 1800s to prove which:
  - Pentagon (5 sides): **yes**. A 17-sided regular polygon: **yes** (Gauss, age 19).
  - 7-sided regular polygon: **no**. Trisecting a general angle: **no**. Squaring
    the circle: **no**.

These are real theorems, not "nobody figured it out yet."

---

# Competition gem

Someone hands you a circle but has **erased the center**. With only compass and
straightedge, find the center.

--

Hint: any chord's perpendicular bisector passes through the center. How many
chords do you need?

---

class: center, middle

# Next up

**Module 2:** tear the corners off any triangle. They always make a straight
line. Why *always*? That is where proof really begins.

Then play: **euclidea.xyz** turns these exact moves into puzzle levels.
