\documentclass[11pt]{article}
\usepackage{geometry}
\geometry{margin=1in}
\usepackage{amssymb}
\usepackage{enumitem}
\usepackage{parskip}
\usepackage{xcolor}
\definecolor{benblue}{HTML}{1769AA}

\newcommand{\drawbox}[1]{%
  \par\vspace{4pt}\noindent
  \fbox{\rule{0pt}{#1}\rule{\dimexpr\linewidth-2\fboxsep-2\fboxrule}{0pt}}%
  \par\vspace{6pt}}

\newcommand{\task}[1]{\par\medskip\noindent\textbf{\color{benblue}#1}\par\nopagebreak}

\title{\vspace{-1.5cm}\textbf{Compass and Straightedge}\\[-2pt]
  \large The oldest game in math}
\author{}
\date{}

\begin{document}
\maketitle
\vspace{-1.2cm}

\noindent\textbf{The one rule.} You may only do two things:
(1) with the straightedge, draw a line through two points;
(2) with the compass, draw a circle from a center and a radius you set on two
points. The ruler's numbers are off limits. It is only an edge.

\section*{Build 1: the perfect triangle}
\begin{enumerate}[leftmargin=*]
  \item Draw a segment, ends $A$ and $B$.
  \item Compass on $A$, pencil on $B$: draw a circle. Same width, compass on $B$:
        draw another.
  \item Call the top crossing $C$. Draw $AC$ and $BC$.
\end{enumerate}
\task{Do it here:}
\drawbox{4.5cm}
\task{Why are all three sides equal, even though you never measured? Write the reason.}
\drawbox{2.2cm}

\section*{Build 2: cut a segment exactly in half}
Compass on $A$ (opened past halfway), arc above and below. Same width on $B$,
arcs crossing. Draw the line through the two crossings. It halves $AB$ \emph{and}
makes a right angle.
\drawbox{4cm}

\section*{Build 3: the six-petal flower}
Draw a circle. \textbf{Keep the compass width fixed.} Step the point around the
circle, marking as you go. Connect the marks.
\task{How many steps to get back to start? Count: \underline{\hspace{2cm}}}
\drawbox{4.5cm}
\task{Why is it \emph{exactly} that number? (Hint: connect the center to two
next-door marks. What kind of triangle is that?)}
\drawbox{2.2cm}

\section*{Push further}
\begin{enumerate}[leftmargin=*]
  \item \textbf{Bisect an angle} with only compass and straightedge.
  \item \textbf{Build a square.}
  \item \textbf{Competition gem.} Someone erased the center of a circle. Find it
        again with compass and straightedge. (Hint: a chord's perpendicular
        bisector passes through the center.)
  \item \textbf{True or false?} You can build each of these with compass and
        straightedge. Guess first, then ask.
        \begin{itemize}
          \item A regular pentagon (5 sides). \hfill T \quad / \quad F
          \item A regular 7-sided polygon. \hfill T \quad / \quad F
          \item Cutting any angle into three equal parts. \hfill T \quad / \quad F
        \end{itemize}
\end{enumerate}

\end{document}
