---
title: "Module 5: Pythagoras"
subtitle: "Proofs you can see"
output:
  xaringan::moon_reader:
    css: [default, default-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = "center", dev = "png",
                      fig.width = 5, fig.height = 5, dpi = 150)
blue   <- "#1769aa"
orange <- "#c2410c"
```

class: center, middle

# Pythagoras

### Proofs you can see

For a right triangle, `a^2 + b^2 = c^2`. You know the formula. Today we *watch*
it become true, three different ways.

---

# The claim, in pictures

.pull-left[
A **right triangle** has one 90-degree corner.

- The two short sides are the **legs**, `a` and `b`.
- The long side across from the corner is the **hypotenuse**, `c`.

Build a square on each side. The claim: the two leg squares, added together,
exactly fill the hypotenuse square.

`a^2 + b^2 = c^2`
]

.pull-right[
```{r squares, fig.width=5, fig.height=5}
par(mar=c(0,0,0,0), pty="s")
a <- 3; b <- 4
plot(NA, xlim=c(-3.5,7.5), ylim=c(-4.5,7.5), asp=1, axes=FALSE, xlab="", ylab="")
# triangle: right angle at origin, legs along axes
A <- c(0,0); B <- c(b,0); C <- c(0,a)
polygon(c(A[1],B[1],C[1]), c(A[2],B[2],C[2]), col="grey90", border=blue, lwd=3)
# square on bottom leg b (downward)
polygon(c(0,b,b,0), c(0,0,-b,-b), col="#cfe3f2", border=blue, lwd=2)
text(b/2, -b/2, expression(b^2), cex=1.3)
# square on left leg a (leftward)
polygon(c(0,0,-a,-a), c(0,a,a,0), col="#cfe3f2", border=blue, lwd=2)
text(-a/2, a/2, expression(a^2), cex=1.3)
# square on hypotenuse (outward)
n <- c(a,b)          # outward normal direction (length c)
B2 <- B + n; C2 <- C + n
polygon(c(B[1],C[1],C2[1],B2[1]), c(B[2],C[2],C2[2],B2[2]),
        col="#f6dcc9", border=orange, lwd=2)
text(mean(c(B[1],C[1],C2[1],B2[1])), mean(c(B[2],C[2],C2[2],B2[2])),
     expression(c^2), cex=1.3)
text(-0.35,-0.3,"a", col=blue, cex=1.2)
text(b/2,0.3,"b", col=blue, cex=1.2)
```
]

---

# The hook

You can recite `a^2 + b^2 = c^2`.

--

But can you **prove** it? Show me *why* it has to be true, for every right
triangle at once, not just the ones you tried?

--

We will do it three ways. Same fact, three pictures. Pick your favorite.

---

# Proof 1: the tilted square

.pull-left[
Big square, side `a + b`. Drop a copy of the triangle into each corner.

The hole they leave is a tilted square of side **c**.

Area two ways:

- One square: `(a+b)^2 = a^2 + 2ab + b^2`
- Pieces: `4 (1/2 ab) + c^2 = 2ab + c^2`
]

.pull-right[
```{r rearr, fig.width=5, fig.height=5}
par(mar=c(0,0,0,0), pty="s")
a <- 3; b <- 4; s <- a+b
plot(NA, xlim=c(-0.4,s+0.4), ylim=c(-0.4,s+0.4), asp=1, axes=FALSE, xlab="", ylab="")
# inner tilted square vertices on the sides
P <- rbind(c(a,0), c(s,a), c(b,s), c(0,b))
# four corner triangles
polygon(c(a,s,s), c(0,0,a), col="#cfe3f2", border=blue, lwd=2)   # bottom-right
polygon(c(s,s,b), c(a,s,s), col="#cfe3f2", border=blue, lwd=2)   # top-right
polygon(c(b,0,0), c(s,s,b), col="#cfe3f2", border=blue, lwd=2)   # top-left
polygon(c(0,a,0), c(b,0,0), col="#cfe3f2", border=blue, lwd=2)   # bottom-left
# inner square
polygon(P[,1], P[,2], col="#f6dcc9", border=orange, lwd=3)
text(s/2, s/2, expression(c^2), cex=1.5)
# outer square
rect(0,0,s,s, border="grey50", lwd=2)
text(a/2,-0.05,"a", col=blue, cex=1.1, pos=1)
text((a+s)/2,-0.05,"b", col=blue, cex=1.1, pos=1)
```
]

---

# Proof 1 finishes itself

The two areas are the same square, so:

```
a^2 + 2ab + b^2  =  2ab + c^2
```

--

Cancel `2ab` from both sides:

```
a^2 + b^2 = c^2
```

--

The `2ab` was just the four triangles, hiding in plain sight. That is the whole
proof.

---

# Proof 2: Garfield's trapezoid

.pull-left[
Found by James Garfield in 1876, **five years before** he became US president.

Two triangles lean into a right **trapezoid**. The middle piece is forced to be a
right triangle with both legs `c`.

- As a trapezoid: `(1/2)(a+b)(a+b)`
- As three triangles: `(1/2)ab + (1/2)ab + (1/2)c^2`
]

.pull-right[
```{r garfield, fig.width=5, fig.height=4.6}
par(mar=c(0,0,0,0), pty="s")
a <- 3; b <- 4
# trapezoid vertices: (0,0),(a+b,0),(a+b,b),(0,a)
plot(NA, xlim=c(-0.6,a+b+0.6), ylim=c(-0.6,b+0.8), asp=1, axes=FALSE, xlab="", ylab="")
P0 <- c(0,0); P1 <- c(a+b,0); P2 <- c(a+b,b); P3 <- c(0,a); Q <- c(b,0)
polygon(c(P0[1],P1[1],P2[1],P3[1]), c(P0[2],P1[2],P2[2],P3[2]),
        col="grey92", border="grey50", lwd=2)
# triangle 1: (0,0),(b,0),(0,a)
polygon(c(0,b,0), c(0,0,a), col="#cfe3f2", border=blue, lwd=2)
# triangle 2: (b,0),(a+b,0),(a+b,b)
polygon(c(b,a+b,a+b), c(0,0,b), col="#cfe3f2", border=blue, lwd=2)
# middle triangle: (b,0),(0,a),(a+b,b)
polygon(c(b,0,a+b), c(0,a,b), col="#f6dcc9", border=orange, lwd=2)
segments(Q[1],Q[2],P3[1],P3[2], col=orange, lwd=2)
segments(Q[1],Q[2],P2[1],P2[2], col=orange, lwd=2)
text(b/2,-0.05,"b", col=blue, cex=1.1, pos=1)
text((b+a+b)/2,-0.05,"a", col=blue, cex=1.1, pos=1)
text(-0.18,a/2,"a", col=blue, cex=1.1, pos=2)
text(a+b+0.18,b/2,"b", col=blue, cex=1.1, pos=4)
text(2.7,2.3,"c", col=orange, cex=1.1)
```
]

---

# Proof 2 finishes itself

Set the two areas equal and multiply by 2:

```
(a+b)^2 = ab + ab + c^2
```

--

```
a^2 + 2ab + b^2 = 2ab + c^2
```

--

```
a^2 + b^2 = c^2
```

A president's doodle lands in the exact same place as Euclid's tilted square.

---

# Proof 3: similar triangles

Drop a line from the right-angle corner straight down onto the hypotenuse. It
splits `c` into two pieces, `p` and `q`, with `p + q = c`.

--

Each small triangle is a **shrunk copy** of the whole one (same three angles).
Matching sides of similar triangles gives:

```
a^2 = c * p          b^2 = c * q
```

--

Add them:

```
a^2 + b^2 = c*p + c*q = c(p + q) = c^2
```

No cutting, no area counting. Two similar triangles and one addition.

---

# Push further: triples

.pull-left[
Whole-number right triangles are **Pythagorean triples**:

`3-4-5`, `5-12-13`, `8-15-17`, `7-24-25` ...

Check one: `5^2 + 12^2 = 25 + 144 = 169 = 13^2`.

There are infinitely many, and there is a **machine** that makes them.
]

.pull-right[
**The triple machine.** Pick whole numbers `m > n`:

`(m^2 - n^2,  2mn,  m^2 + n^2)`

is always a triple.

- `m=2, n=1` gives `(3, 4, 5)`
- `m=3, n=2` gives `(5, 12, 13)`
- `m=4, n=1` gives `(15, 8, 17)`

It never fails, because
`(m^2-n^2)^2 + (2mn)^2 = (m^2+n^2)^2`.
]

---

# Push further: where it breaks

- **Not-right triangles.** If the angle is not 90 degrees, `a^2 + b^2` misses
  `c^2`. The full fix is the **law of cosines**: `c^2 = a^2 + b^2 - 2ab cos C`.
  At `C = 90`, `cos C = 0`, and Pythagoras drops right back out.

--

- **3D.** A box with edges `a, b, c` has space diagonal `d` with
  `d^2 = a^2 + b^2 + c^2`. It is just Pythagoras used twice: once on the floor,
  once standing up.

---

# Competition gem: the converse

The theorem says: right angle `=>` `a^2 + b^2 = c^2`.

--

The **converse** runs the arrow backwards: if `a^2 + b^2 = c^2`, then the angle
opposite `c` *must* be a right angle.

--

That is the real reason a knotted 3-4-5 rope makes a perfect right corner, and
why builders have trusted it for thousands of years.

--

**Prove it.** Build a fresh right triangle with legs `a` and `b`; its hypotenuse
is `c`, so it has the same three sides as yours, so the triangles are congruent,
so your angle is right too.

---

class: center, middle

# The finish line

That was the last build-and-prove module. You have compass constructions, the
angle sum, circle angles, tilings, and now Pythagoras, each ending in a proof.

**Next: compete.** A set of BMC-style problem sheets (angle chasing, cyclic
quadrilaterals, constructions) that braid Modules 01 through 05 together, once
Parts 1 and 2 feel solid.
