---
title: "Module 4: Tilings"
subtitle: "Why only three shapes fill the floor"
output:
  xaringan::moon_reader:
    css: [default, default-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = "center", dev = "png",
                      fig.width = 5, fig.height = 5, dpi = 150)
blue   <- "#1769aa"
orange <- "#c2410c"
lblue  <- "#cfe3f3"
lred   <- "#fdd9d0"

hexpoly <- function(cx, cy, R, fill, border) {
  a <- ((0:5)*60 + 90) * pi/180
  polygon(cx + R*cos(a), cy + R*sin(a), col=fill, border=border, lwd=1.8)
}

draw_tri_tiling <- function(main="") {
  h <- sqrt(3)/2; ny <- 3
  par(mar=c(0.4,0.4,2,0.4))
  plot(NA, xlim=c(0.5, 4.0), ylim=c(0, ny*h), asp=1, axes=FALSE,
       xlab="", ylab="", main=main, cex.main=1.5)
  for (r in 0:(ny-1)) {
    y0 <- r*h; y1 <- y0+h; off <- (r %% 2)*0.5
    for (i in -1:4) {
      bx <- i + off
      polygon(c(bx, bx+1, bx+0.5), c(y0,y0,y1), col=lblue, border=blue, lwd=1.8)
      polygon(c(bx+0.5, bx+1.5, bx+1), c(y1,y1,y0), col=lblue, border=blue, lwd=1.8)
    }
  }
}

draw_sq_tiling <- function(main="") {
  n <- 4
  par(mar=c(0.4,0.4,2,0.4))
  plot(NA, xlim=c(0,n), ylim=c(0,n), asp=1, axes=FALSE,
       xlab="", ylab="", main=main, cex.main=1.5)
  for (i in 0:(n-1)) for (j in 0:(n-1))
    rect(i,j,i+1,j+1, col=lblue, border=blue, lwd=1.8)
}

draw_hex_tiling <- function(main="") {
  R <- 1; w <- sqrt(3)*R; nc <- 3; nr <- 3
  par(mar=c(0.4,0.4,2,0.4))
  plot(NA, xlim=c(0, nc*w), ylim=c(0, nr*1.5*R), asp=1, axes=FALSE,
       xlab="", ylab="", main=main, cex.main=1.5)
  for (r in -1:(nr+1)) for (c in -1:(nc+1)) {
    cx <- c*w + (r %% 2)*(w/2); cy <- r*1.5*R
    hexpoly(cx, cy, R, lblue, blue)
  }
}

draw_vertex360 <- function() {
  par(mar=c(0,0,0,0))
  plot(NA, xlim=c(-2.4,2.4), ylim=c(-2.4,2.4), asp=1, axes=FALSE, xlab="", ylab="")
  R <- 1.4
  for (phi in c(150,270,30)) {
    pc <- R*c(cos(phi*pi/180), sin(phi*pi/180))
    a  <- (phi+180 + 60*(0:5))*pi/180
    polygon(pc[1]+R*cos(a), pc[2]+R*sin(a), col=lblue, border=blue, lwd=2.2)
    text(0.55*cos(phi*pi/180), 0.55*sin(phi*pi/180), "120", cex=1.2, col=blue)
  }
  points(0,0, pch=19, cex=1.4)
}

draw_pentagon_gap <- function() {
  par(mar=c(0,0,0,0))
  plot(NA, xlim=c(-2.4,2.4), ylim=c(-2.0,2.6), asp=1, axes=FALSE, xlab="", ylab="")
  Rp <- 1.3; s <- 2*Rp*sin(36*pi/180)
  # gap wedge between 72 and 108 degrees
  g <- c(72,108)*pi/180
  polygon(c(0, s*cos(g[1]), s*cos(g[2])), c(0, s*sin(g[1]), s*sin(g[2])),
          col=lred, border=NA)
  for (phi in c(162,270,18)) {
    pc <- Rp*c(cos(phi*pi/180), sin(phi*pi/180))
    a  <- (phi+180 + 72*(0:4))*pi/180
    polygon(pc[1]+Rp*cos(a), pc[2]+Rp*sin(a), col=lblue, border=blue, lwd=2.2)
    text(0.5*cos(phi*pi/180), 0.5*sin(phi*pi/180), "108", cex=1.1, col=blue)
  }
  points(0,0, pch=19, cex=1.4)
  text(0, 0.95, "36 gap", cex=1.2, col=orange, font=2)
}
```

class: center, middle

# Tilings

### Why only three shapes fill the floor

A tiled floor, a honeycomb. The pieces cover everything with no gaps and no
overlaps. If every tile is the same regular shape, which shapes can do it?

---

# The setup

A **tiling** covers the plane with tiles: **no gaps, no overlaps.**

--

Today's rule: every tile is the **same regular polygon** (all sides equal, all
corners equal).

--

Which regular shapes can do it? Triangle? Square? Pentagon? Hexagon? Make a
guess for each before we test them.

---

# The hook

Try ringing a single point with copies of one shape, corner to corner.

--

- Equilateral **triangles**: they close up. Works.
- **Squares**: four to a corner, like graph paper. Works.
- **Hexagons**: lock together like a honeycomb. Works.

--

- Regular **pentagons**: push them together as tight as you can...

--

...a thin **wedge of gap** is always left over. Pentagons **never** close up.
Why exactly those three shapes, and not the pentagon?

---

# Three that work

```{r tilings3, fig.width=9, fig.height=3.2}
par(mfrow=c(1,3))
draw_tri_tiling("Triangles")
draw_sq_tiling("Squares")
draw_hex_tiling("Hexagons")
```

Triangle, square, hexagon. That is the whole list. The reason is a single idea
about what happens at one **meeting-point**.

---

# The key question: one vertex

.pull-left[
Zoom in on any point where corners meet. To leave **no gap and no overlap**, the
corners crowding around that point must sweep the **whole turn**:

$$\text{they add to exactly } 360^\circ.$$

Three hexagon corners do it: `120 + 120 + 120 = 360`.

So the question becomes: **whose corner angle divides 360 evenly?**
]

.pull-right[
```{r v360, fig.width=4.6, fig.height=4.6}
draw_vertex360()
```
]

---

# The corner angle of a regular n-gon

From **Module 02**: the angles inside an n-gon add to `(n - 2) * 180`.

--

A *regular* n-gon shares that total evenly over its n corners:

$$\text{corner angle} = \frac{(n-2)\cdot 180}{n}$$

--

| n | 3 | 4 | 5 | 6 | 7 |
|---|---|---|---|---|---|
| corner | 60 | 90 | 108 | 120 | 128.57... |

For `k` copies to fit snugly around a point: `k * (corner) = 360`, so the corner
must **divide 360 evenly**.

---

# Why must it be exactly those three?

Check which corner angles divide 360:

--

- triangle, **60**: `360 / 60 = 6`. Six fit. **Works.**
- square, **90**: `360 / 90 = 4`. **Works.**
- hexagon, **120**: `360 / 120 = 3`. **Works.**

--

- pentagon, **108**: `360 / 108 = 3.33...`. Three give `324` (a `36` gap), four
  give `432` (overlap). **No whole number fits.**

--

And **nothing past the hexagon** can work: for `n >= 7` the corner is between
`120` and `180`, so only **two** could meet, and two angles under `180` never
reach `360`. Only `60, 90, 120` divide 360. **Exactly three tilings.**

---

# The pentagon's gap, up close

.pull-left[
Three regular pentagons around a point:

$$108 + 108 + 108 = 324.$$

That leaves a `360 - 324 = 36^\circ` wedge. A fourth pentagon needs `108` but
only `36` is left, so it overlaps.

There is no whole number of `108`s that hits `360`. The gap is **forced**, for
the same reason the three winners are forced.
]

.pull-right[
```{r pgap, fig.width=4.6, fig.height=4.6}
draw_pentagon_gap()
```
]

---

# Push further: mix the shapes

Drop "all tiles the same." Keep two rules: corners at each vertex still total
**360**, and every vertex looks the same. These are the **semiregular** tilings:

--

- **4.8.8**: square + two octagons. `90 + 135 + 135 = 360`. (the "stop sign")
- **3.6.3.6**: `60 + 120 + 60 + 120 = 360`.
- **3.3.3.3.6**: four triangles + a hexagon. `60*4 + 120 = 360`.
- **3.4.6.4**: `60 + 90 + 120 + 90 = 360`.

--

There are exactly **eight** of them. Finding all eight is a great afternoon.

---

# Competition gem

**The search BMC loves:** which multisets of regular-polygon corner angles add
to exactly **360** at one point?

--

It is a *finite* hunt: a big-angle polygon leaves little room. For instance
`3.4.4.6`: `60 + 90 + 90 + 120 = 360`. And `3.12.12`: `60 + 150 + 150 = 360`.

--

The twist: **a valid vertex need not tile the whole plane.** `3.4.4.6` passes the
angle test at one point but cannot grow outward with every vertex matching.

So the real problem has two parts: (1) list every angle-multiset summing to 360,
then (2) decide which actually tile. Part 1 is arithmetic; part 2 is geometry.

---

class: center, middle

# Next up

**Module 5:** Pythagoras. We prove `a^2 + b^2 = c^2` with pictures you *watch*,
not formulas you memorize: rearrangement, the Garfield trapezoid, similar
triangles.

And a bridge to Isa's room: her pattern blocks find this exact pentagon gap by
hand, with no angles at all.
