---
title: "Module 2: Why every triangle adds up to 180"
subtitle: "Where proof begins"
output:
  xaringan::moon_reader:
    css: [default, default-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = "center", dev = "png",
                      fig.width = 5, fig.height = 5, dpi = 150)
blue   <- "#1769aa"
orange <- "#c2410c"

# small helper: draw an angle arc at (cx,cy) from a1 to a2 degrees
arc <- function(cx, cy, r, a1, a2, col, lwd = 3) {
  a <- seq(a1, a2, length.out = 60) * pi / 180
  lines(cx + r * cos(a), cy + r * sin(a), col = col, lwd = lwd)
}
```

class: center, middle

# Why every triangle adds up to 180

### Where proof begins

Draw any triangle you like. Tear off its three corners. They always snap
together into one straight line. Always. Why *always*?

---

# The experiment

.pull-left[
1. Draw three triangles that look nothing alike: a long skinny one, a fat one, a
   pointy one.
2. Shade the three corners of each so you can tell them apart.
3. **Tear** (do not cut) the three corners off.
4. Slide the torn corners together, points touching.
]

.pull-right[
```{r tear, fig.width=5, fig.height=4.2}
par(mar = c(0,0,0,0), pty = "s")
plot(NA, xlim = c(-1.4, 1.4), ylim = c(-0.4, 1.5), asp = 1, axes = FALSE,
     xlab = "", ylab = "")
# straight base line
segments(-1.3, 0, 1.3, 0, col = "grey50", lwd = 2)
# three torn wedges meeting at the origin, filling the half-plane
polygon(c(0, -1.3, -0.45), c(0, 0, 1.25), col = "#cfe2f3", border = blue, lwd = 2)
polygon(c(0, -0.45, 0.45), c(0, 1.25, 1.25), col = "#fde9d9", border = orange, lwd = 2)
polygon(c(0, 0.45, 1.3), c(0, 1.25, 0), col = "#d9ead3", border = "#38761d", lwd = 2)
points(0, 0, pch = 19, cex = 1.2)
text(-0.7, 0.45, "1", cex = 1.5)
text(0, 0.8, "2", cex = 1.5)
text(0.7, 0.45, "3", cex = 1.5)
text(0, -0.28, "a straight line = 180", cex = 1.1, col = "grey30")
```
]

The three corners fill exactly a straight angle. A straight line is **180
degrees**, so the three angles add to 180.

---

# The honest question

You tried it three times. Maybe you got lucky three times.

--

How do you know triangle number **4,000,000**, the one nobody has drawn yet,
also closes up flat?

--

You cannot tear your way through infinitely many triangles. Trying more will
never settle it.

--

We need an **argument** that covers every triangle at once. That is a *proof*,
and this is where it starts.

---

# First, a smaller fact: the "Z" rule

.pull-left[
Two **parallel** lines (use both edges of your ruler). One slanted line crossing
both makes a **Z**.

The two angles tucked inside the Z, one at each parallel line, on opposite sides
of the slant, are **equal**.

This is the **alternate angles** rule. It is *the* fact about parallel lines.
]

.pull-right[
```{r zrule, fig.width=5, fig.height=4.4}
par(mar = c(0,0,0,0))
plot(NA, xlim = c(-1.2, 3.2), ylim = c(-0.5, 2.5), asp = 1, axes = FALSE,
     xlab = "", ylab = "")
# two parallel horizontal lines
segments(-1, 2, 3, 2, col = "grey50", lwd = 2)
segments(-1, 0, 3, 0, col = "grey50", lwd = 2)
# transversal through B=(0.7,0) and T=(2.3,2)
B <- c(0.7, 0); T <- c(2.3, 2)
segments(B[1] - 0.55, B[2] - 0.7, T[1] + 0.35, T[2] + 0.45, col = blue, lwd = 3)
# alternate interior angles
arc(T[1], T[2], 0.55, 180, 231.3, orange, lwd = 4)
arc(B[1], B[2], 0.55, 0, 51.3, orange, lwd = 4)
points(B[1], B[2], pch = 19); points(T[1], T[2], pch = 19)
text(T[1] - 0.85, T[2] - 0.35, "a", cex = 1.5, col = orange)
text(B[1] + 0.85, B[2] + 0.3, "a", cex = 1.5, col = orange)
```
]

Slide one marked angle along the slant until it lands on the other: they match.

---

# Now the proof

.pull-left[
Take any triangle **ABC**. Through the top vertex **A**, draw a line **parallel
to the bottom side BC**.

Three angles sit along that straight line at A: **x**, the apex angle **A**, and
**y**. They lie on a straight line, so

$$x + A + y = 180.$$
]

.pull-right[
```{r proof, fig.width=5, fig.height=4.4}
par(mar = c(0,0,0,0), pty = "s")
A <- c(0.7, 1.6); Bp <- c(0, 0); Cp <- c(2, 0)
plot(NA, xlim = c(-1.0, 2.6), ylim = c(-0.4, 2.1), asp = 1, axes = FALSE,
     xlab = "", ylab = "")
# line through A parallel to BC (horizontal)
segments(-0.9, A[2], 2.5, A[2], col = "grey50", lwd = 2)
# triangle
polygon(c(A[1], Bp[1], Cp[1]), c(A[2], Bp[2], Cp[2]), border = blue, lwd = 3)
points(rbind(A, Bp, Cp), pch = 19, cex = 1.1)
text(A[1] - 0.18, A[2] + 0.16, "A", cex = 1.3)
text(Bp[1] - 0.18, Bp[2] - 0.05, "B", cex = 1.3, col = orange)
text(Cp[1] + 0.18, Cp[2] - 0.05, "C", cex = 1.3, col = "#38761d")
# the three angles along the line at A
text(A[1] - 0.42, A[2] - 0.12, "x", cex = 1.3, col = orange)
text(A[1] + 0.02, A[2] - 0.26, "A", cex = 1.2, col = blue)
text(A[1] + 0.45, A[2] - 0.12, "y", cex = 1.3, col = "#38761d")
```
]

--

.pull-right[
Now the **Z rule**, twice:

- **x = B** (alternate angles across AB)
- **y = C** (alternate angles across AC)
]

---

# The leap

Start from the straight line at A:

$$x + A + y = 180.$$

--

Swap in `x = B` and `y = C`:

$$B + A + C = 180.$$

--

We never used the *shape* of the triangle, only that the helper line through A
was parallel to BC, which you can **always** draw.

--

So it holds for **every** triangle at once. That is the jump from "I tore three
and they worked" to "it is *impossible* for one to fail."

---

# More sides? Split into triangles

.pull-left[
Any polygon cuts into triangles from one corner.

- 4 sides: 2 triangles, $2 \times 180 = 360$.
- 5 sides: 3 triangles, $3 \times 180 = 540$.

An **n**-sided polygon makes $n - 2$ triangles:

$$(n - 2) \times 180.$$
]

.pull-right[
```{r poly, fig.width=5, fig.height=4.6}
par(mar = c(0,0,0,0), pty = "s")
k <- 0:4
th <- (90 + 72 * k) * pi / 180
px <- cos(th); py <- sin(th)
plot(NA, xlim = c(-1.25, 1.25), ylim = c(-1.25, 1.25), asp = 1, axes = FALSE,
     xlab = "", ylab = "")
polygon(px, py, border = blue, lwd = 3)
# diagonals from vertex 1 splitting into 3 triangles
segments(px[1], py[1], px[3], py[3], col = orange, lwd = 2)
segments(px[1], py[1], px[4], py[4], col = orange, lwd = 2)
points(px, py, pch = 19, cex = 1.1)
```
]

Have Ben derive the formula himself by drawing the cuts.

---

# The walking-turtle trick

Imagine walking all the way around the outside of any polygon and back to your
start, facing your original direction.

--

The total amount you **turned** is exactly **one full turn = 360 degrees**, no
matter how many sides.

--

Those turns are the **exterior angles**. So the exterior angles of any convex
polygon add to **360**, full stop.

--

Often slicker than the interior-angle formula, and a real math-circle way of
thinking: don't sum the corners, *track the turning*.

---

# Competition gem: the five-pointed star

.pull-left[
Draw a five-pointed star in one stroke. It has five sharp points.

**What do the five point-angles add up to?**

The answer is a clean **180 degrees**, and it does *not* depend on how lopsided
your star is.
]

.pull-right[
```{r star, fig.width=5, fig.height=4.8}
par(mar = c(0,0,0,0), pty = "s")
k <- 0:4
th <- (90 + 72 * k) * pi / 180
px <- cos(th); py <- sin(th)
ord <- c(1, 3, 5, 2, 4, 1)
plot(NA, xlim = c(-1.25, 1.25), ylim = c(-1.3, 1.25), asp = 1, axes = FALSE,
     xlab = "", ylab = "")
polygon(px[ord], py[ord], border = blue, lwd = 3)
points(px, py, pch = 19, cex = 1.4, col = orange)
```
]

--

.pull-right[
Hint: each point is the tip of a little triangle. Chase the angles using the 180
rule and the straight-line rule.
]

---

class: center, middle

# Next up

**Module 3:** Circle secrets, the inscribed angle.

An angle that pins itself to a circle behaves in a way that looks like magic,
and the 180 rule you just proved is exactly the key that unlocks it.
